Well-quasi-orderings on word languages

### 2025-10-15 src/appendix.tex
Move some stuff to appendix

>>>>

\begin{lemma}
	\label{automatic-uur:lem}
	\proofref{automatic-uur:lem}
	Given an \kl{automatic sequence} $w \in \Sigma^{\Nat}$, one can decide
	whether it is \kl{ultimately uniformly recurrent}.
\end{lemma}

\begin{proofof}{automatic-uur:lem}
	We can rewrite this as a question on the \kl{automatic sequence} $w$
	as follows:
	\begin{align*}
		&\exists N_0,                   &   \text{ultimately} \\
		&\forall i_s \geq N_0,          &   \text{for every infix (start) } u \\
		&\forall i_e > i_s,             &   \text{for every infix (end) }   u \\
		&\exists k \geq 1,              &   \text{there exists a bound} \\
		&\forall j_s \geq N_0,          &   \text{for every other infix (start) } v \\
		&\forall j_e \geq j_s + k,      &   \text{of size at least $k$} \\
		&\exists l \geq 0,              &   \text{there exists a position in } v \\
		&\forall 0 \leq m < i_e - i_s,  &   \text{where } u \text{ can be found} \\
		&j_s + m + l < j_e \land
		w(i_s + m) = w(j_s + m + l) \quad .
	\end{align*}
	Because $w$ is computable by a finite automaton, one can reduce the above
	formula to a regular language, for which it suffices to check emptiness, which
	is decidable.
\end{proofof}

### 2025-10-15 src/infixes.tex
Move some stuff to appendix

\AP A first model that one can use to represent infinite words is the model of
\intro{automatic sequences}. In this case, the infinite word $w$ is described
by a finite state automaton, that can compute the $i$-th letter of the word $w$
given as input the number $i$ written in some base $b \in \Nat$. An example of
such a sequence is the \kl{Thue-Morse sequence} that can be described by a
finite automaton using a binary representation of the indices. The good
algorithmic properties of \kl{automatic sequences} come from the fact that a
Presburger definable property that uses letters of the sequence can be
(trivially) translated into a finite automaton that reads the base $b$
representation of the free variables (that are indices of the sequence). We
briefly explain in the following lemma the folklore result that one can decide
if an \kl{automatic sequence} is \kl{ultimately uniformly recurrent}.

\begin{lemma}
  \label{automatic-uur:lem}
    Given an \kl{automatic sequence} $w \in \Sigma^{\Nat}$, one can decide
    whether it is \kl{ultimately uniformly recurrent}.
\end{lemma}
\begin{proof}
    We can rewrite this as a question on the \kl{automatic sequence} $w$
    as follows:
    \begin{align*}
        &\exists N_0,                   &   \text{ultimately} \\
        &\forall i_s \geq N_0,          &   \text{for every infix (start) } u \\
        &\forall i_e > i_s,             &   \text{for every infix (end) }   u \\
        &\exists k \geq 1,              &   \text{there exists a bound} \\
        &\forall j_s \geq N_0,          &   \text{for every other infix (start) } v \\
        &\forall j_e \geq j_s + k,      &   \text{of size at least $k$} \\
        &\exists l \geq 0,              &   \text{there exists a position in } v \\
        &\forall 0 \leq m < i_e - i_s,  &   \text{where } u \text{ can be found} \\
        &j_s + m + l < j_e \land
        w(i_s + m) = w(j_s + m + l) \quad .
    \end{align*}
    Because $w$ is computable by a finite automaton, one can reduce the above
    formula to a regular language, for which it suffices to check emptiness, which
    is decidable.
\end{proof}

Based on \cref{automatic-uur:lem}, we can now prove the following theorem.
\begin{theorem}
    \label{automatic-wqo:thm}
    Given a \kl{sequence representation} of a language $L \subseteq
    \Sigma^*$ where all infinite words are \kl{automatic sequences}, one can
    decide whether $L$ is \kl{well-quasi-ordered} for the \kl{infix relation}.
\end{theorem}
\begin{proof}
    It is easy to see that $L$ is \kl{well-quasi-ordered} for the \kl{infix
    relation} if and only if for every triple $(w_i^-, a_i, w_i^+)$ in the
    \kl{sequence representation} of $L$, the (potentially bi-infinite) word
    $\mathsf{reversed}(w_i^-) a_i w_i^+$ defines a \kl{well-quasi-ordered}
    language. By \cref{from-bi-to-single:lem}, this is the case if and only if
    both $w_i^-$ and $w_i^+$ are \kl{ultimately uniformly recurrent}. 
    Since one can decide whether an \kl{automatic sequence} is
    \kl{ultimately uniformly recurrent} using \cref{automatic-uur:lem},
    we conclude the proof.
\end{proof}

>>>>

\AP A first model that one can use to represent infinite words is the model of
\intro{automatic sequences}. In this case, the infinite word $w$ is described
by a finite state automaton, that can compute the $i$-th letter of the word $w$
given as input the number $i$ written in some base $b \in \Nat$. An example of
such a sequence is the \kl{Thue-Morse sequence} that can be described by a
finite automaton using a binary representation of the indices. The good
algorithmic properties of \kl{automatic sequences} come from the fact that a
Presburger definable property that uses letters of the sequence can be
(trivially) translated into a finite automaton that reads the base $b$
representation of the free variables (that are indices of the sequence). It is folklore that one can decide if an \kl{automatic sequence} is \kl{ultimately uniformly recurrent}. A proof can be found in the appendix at \cref{automatic-uur:lem}. Based on this, we now prove:

\begin{theorem}
    \label{automatic-wqo:thm}
    Given a \kl{sequence representation} of a language $L \subseteq
    \Sigma^*$ where all infinite words are \kl{automatic sequences}, one can
    decide whether $L$ is \kl{well-quasi-ordered} for the \kl{infix relation}.
\end{theorem}
\begin{proof}
    It is easy to see that $L$ is \kl{well-quasi-ordered} for the \kl{infix
    relation} if and only if for every triple $(w_i^-, a_i, w_i^+)$ in the
    \kl{sequence representation} of $L$, the (potentially bi-infinite) word
    $\mathsf{reversed}(w_i^-) a_i w_i^+$ defines a \kl{well-quasi-ordered}
    language. By \cref{from-bi-to-single:lem}, this is the case if and only if
    both $w_i^-$ and $w_i^+$ are \kl{ultimately uniformly recurrent}. 
    Since one can decide whether an \kl{automatic sequence} is
    \kl{ultimately uniformly recurrent} using \cref{automatic-uur:lem},
    we conclude the proof.
\end{proof}
